Detecting Stylistic Variation in Pop Production

### 2023-12-04 outputs/paper/paper.qmd
thanks
Minor tweaks throughout

---
title: "Detecting Stylistic Variation in Pop Production"
author: 
  - name: Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop. We thank Rohan Alexander, Carl Wilson and Inessa De Angelis for valuable suggestions. For any comments or suggestions, please contact michaela.drouillard@mail.utoronto.ca"
date: "today"
date-format: "long"
abstract: "This study investigates the feasibility of predicting whether a song is produced by Jack Antonoff using Spotify's audio feature API. We developed two models: a logistic regression model and a random forest model. The logistic regression model, focusing on Antonoff's collaborators' discographies, demonstrates an accuracy of 76%, precision of 39%, and a recall of 76% when predicting whether a track has an Antonoff credit. The random forest model, which includes discographies of other notable pop producers, shows an accuracy of 61% when predicting tracks' producers, comparing favorably to a No Information Rate of 43%. A significant finding is that 'danceability', a metric developed from human-tagged data, emerges as a crucial predictor in both models. This research contributes to the broader understanding of computational analysis in cultural contexts, illustrating how a blend of human judgment and algorithmic processing can approximate musical styles."
format: pdf
number-sections: true
bibliography: references.bib
csl: modern-language-association.csl
number-depth: 2
---

@@

library(tidyverse)
library(lubridate)
library(tidyr)
library(dplyr)
library(tidymodels)
library(randomForest)
library(ranger)
library(caret)
library(themis)
library(yardstick)

```

@@

It is widely acknowledged that style recognition in music is difficult. Style, which is hard enough to define without building computational representations of it, is often constructed using lower-level perceptual features such as a pitch and tempo. We are interested in understanding how to create an approximation for production style based on a data set built using the Spotify audio features API, which includes both crowdsourced features and features derived from machine listening techniques. To do this, we explore different audio feature metrics and identify which were the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We obtained tracks from six artists with extensive discographies who have collaborated with Antonoff, both before and during their collaboration. We also obtained tracks from similarly prolific pop producers: Joel Little, Ariel Rechstaid, Max Martin, Greg Kurstin, Paul Epworth and Rick Nowels. Our data set contains 3738 tracks in total, which were split into a training and test sets.

We built two models to explore the data. The first model, a logistic regression, is employed to observe any patterns in audio feature scores across Antonoff's collaborations with artists. The visualizations and analysis offer a closer look at Antonoff's oeuvre, along with a portrait of the trends in female singer-songwriter contemporary pop (according to Spotify's API). It draws on a dataset of artists who have collaborated with Antonoff (1954 tracks total), with a binary variable to tag whether the song was an Antonoff production. The training set included 1465 tracks, and the testing set included 489 tracks.

The second model, a Random Forest (RF), trains on a data set that includes the discographies of other prominent pop producers to classify producers. This expands upon and complements the results of the logistic regression -- where a logistic regression predicting an Antonoff-produced track in an artist's discography might actually be picking up on an artist releasing pop music, training a Random Forest on a bigger dataset with more producers captures both non-linear relationships in the data and accounts for intra-genre differences. In obtaining the strongest predictor variables in the Random Forest, we can better understand how it distinguishes between different pop producers' styles.

@@

In an interview with Spotify principal engineer Glen McDonald, we learned that the danceability and valence features were built using crowdsourced methods. With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making also shape the recommendation systems that they build (@citeSeaver) -- our results suggest that it's possible to create representations of style by combining features derived from both automated audio analysis and more crowdsourced, subjective contributions. This research adds to a more informed understanding of how style and tastes can been algorithmically rendered using Spotify's audio features.

The remainder of this paper is structured as follows: @sec-data details the data curation and extraction processes, including documentation on features from Spotify's audio features data set. @sec-model specifies the models used and @sec-results details their performance and most important predictors. @sec-discussion situates the study in the broader context of computational stylistics and recommendation systems, detailing how we can conceptualize "style" given the results and limitations of our study. We provide background on the Spotify features' provenance based on an interview we conducted with Glen McDonald, a principal engineer at Spotify.

@@

We collected data from the Spotify API, which we accessed using the `spotifyr` package in R [@citeSpotify, @citeR]. For our first dataset, we acquired audio feature data on the complete Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Van Etten, and Mitski were included to represent other contemporary pop artists with overlapping fan bases who have never collaborated with Antonoff. Of the artists who did collaborate with Antonoff, we chose artists who have produced multiple albums, and at least one album with Antonoff. This dataset contains 1954 tracks in total.

For our second dataset, we acquired the discographies of comparable contemporary pop producers' by scraping their respective discography Wikipedia pages. Tracks were included if producers were either the solo producer, the primary producer, or a co-producer (a limitation of this study being that we can not know or truly parse who takes ownership for what during creative collaborations, especially when our audio features use one number to describe an entire song). The producers include Epworth, Rechstaid, Max Martin, Nowels, Joel Little, and Greg Kustin. These producers were chosen based both on their popularity, and because some have collaborated with the same artists as Antonoff. We combined these tracks with the first datasets tracks, for a total of 3738 tracks from 828 different artists[^1].

@@

We created a binary variable, `is_antonoff`, which contains 1 if Jack Antonoff produced or co-produced the song, and 0 if it was produced by somebody else. The information underpinning this discography was drawn from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

```{r}
#| label: tbl-producercounts
#| tbl-cap: Counts of Tracks Per Producer in Original Dataset
#| echo: false
#| warning: false
lr_data<- read_csv(here::here("inputs/data/lr_data.csv"))

library(kableExtra)

@@

tbl |>
  kable(col.names = c("Producer", "Number of Tracks"),
  digits = 1,
  booktabs = TRUE,
  linesep = "")

```

@@

Spotify provides variables for each song, some of which were developed by them, which enables comparisons across artists. The final audio features in our data set are: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

@fig-means depicts the range of scores for the tracks across all variables except for tempo, mode, and loudness, which are measured using different metrics. Spotify documents the `get_artist_audio_features` variables as follows (@citeSpotifyDocumentation):

> **acousticness**: A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.
>
> **danceability**: Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.
>
> **energy**: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.
>
> **instrumentalness**: Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0.
>
> **speechiness:** Speechiness detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks.
>
> **key**: The key the track is in. Integers map to pitches using standard [Pitch Class notation](https://en.wikipedia.org/wiki/Pitch_class). E.g. 0 = C, 1 = C♯/D♭, 2 = D, and so on. If no key was detected, the value is -1.
>
> **liveness**: Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.
>
> **loudness:** The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.
>
> **mode:** Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0.
>
> **tempo:** The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.
>
> **valence:** A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).

```{r, fig.width= 8, fig.height=6}
#| label: fig-means
#| fig-cap: Comparing the Range and Mean Values of Each Variable From the Spotify Audio Features API for our Full Data Set
#| echo: false
#| warning: false

lr_data %>%
  select(-producer, -tempo, -loudness, -mode) %>%
  pivot_longer(cols = -is_antonoff, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = feature, y = value, fill = factor(is_antonoff))) +
  geom_boxplot(alpha = 0.6, position = position_dodge(width = 0.8)) +
  scale_fill_manual(values = c("#B0C4DE", "#9ACD32"), labels = c("No", "Yes")) +
  theme_minimal() +
  labs(fill = "Antonoff", x = "Feature", y = "Value") + 
  coord_flip()

```

By analyzing the cumulative distribution functions of various audio features, particularly when categorizing songs based on whether they were produced by Antonoff (@fig-cdf), distinct patterns emerge. Notably, while attributes such as liveness, instrumentalness, and speechiness exhibit relative homogeneity, there are slight disparities in features like danceability, energy, loudness, acousticness, and valence.

```{r, fig.width= 8, fig.height=6}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable From the Spotify Audio Features API for our Full Data Set
#| echo: false
#| warning: false
av<- read_csv(here::here("inputs/data/antonoff_viz.csv"))

av$is_antonoff <- ifelse(av$producer == "antonoff", 1, 0)
av$is_antonoff <- as.factor(av$is_antonoff)
levels(av$is_antonoff) <- c("No", "Yes") 
 av |>
   pivot_longer(
     cols = c(
       danceability,
       energy,
       loudness,
       mode,
       speechiness,
       acousticness,
       instrumentalness,
       liveness,
       valence,
       tempo
     ),
     names_to = "name",
     values_to = "value"
   ) |>
   ggplot(aes(x = value, color = is_antonoff)) +
   stat_ecdf() +
   facet_wrap(vars(name),
              nrow = 3,
              ncol = 4,
              scales = "free") +
   theme_minimal() +
   labs(x = "Value",
        y = "Proportion",
        color = "Antonoff")

```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-energy
#| fig-cap: Energy Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av <- av |>
  mutate(is_antonoff= factor(is_antonoff))

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-danceability
#| fig-cap: Danceability Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = danceability, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Danceability") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-valence
#| fig-cap: Valence Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = valence, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Valence") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-acousticness
#| fig-cap: Acousticness Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = acousticness, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Acousticness") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-instrumentalness
#| fig-cap: Instrumentalness Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = instrumentalness, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Instrumentalness") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-tempo
#| fig-cap: Tempo Scores For Contemporary Pop Artists Over Time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = tempo, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Tempo") +
  facet_wrap( ~ artist_name, ncol = 3)
```

@@

To be clear, we do not suggest that Antonoff's involvement is the cause for the change in artist's sound. Arguments of "Antonoffication" risk removing agency and artistic credit from (female) artists, and attributing both their critical success and failures to Antonoff (@citeWilson). In an industry primarily led by male producers, and using features so limited in their ability to describe anything about the process of creation, it would not be appropriate the use these trends to suggest that Antonoff's involvement is the cause behind changes in sound. The purpose of these visualizations is to observe the distribution of variables across artists, over time, with a focus on Antonoff.

This exploratory data analysis sets the stage for a) employing logistic regression to quantify the likelihood of Antonoff's involvement in a song based on these audio features, b) classifying tracks produced by a wider range of producers using a Random Forest, c) identifying the most influential variables in this classification to better understand how style can be inferred from the provided data set features.

@@

The aim of the logistic regression is to determine whether we can predict Antonoff's influence on artists' sound. We select a total of 1465 tracks from the dataset from 12 artists: Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Van Etten, Mitski, and Bleachers. Marina and the Diamonds, Maggie Rogers, Sharon Van Etten and Mitski have not collaborated with Antonoff, but their tracks were added to contribute more data from the same taste pool to provide more context for the model.

@@

$$
\begin{aligned}
Pr(y_i = 1) = & \text{logit}^{-1}(β_0 + β_1 \times \text{danceability} + β_2 \times \text{energy} \\
& + β_3 \times \text{loudness} + β_4 \times \text{mode} + β_5 \times \text{speechiness} \\
& + β_6 \times \text{acousticness} + β_7 \times \text{instrumentalness} \\
& + β_8 \times \text{liveness} + β_9 \times \text{valence} + β_{10} \times \text{tempo})
\end{aligned}
$$

where $\text{Pr}(y_i=1)$ is the probability of a song being produced by Antonoff, $\beta_0$ is the intercept, and $\beta_1, \beta_2, \beta_3, \beta_4, \beta_5, \beta_6, \beta_7, \beta_8, \beta_9, \beta_{10}$ are the coefficients corresponding to the 10 predictor variables.

After some minor pre-processing using the `tidyverse` package [@citetidyverse] in R [@citeR], we followed Julia Silge's downsampling methodology to build a model that accounts for imbalanced classes (@citeSilge). We downsampled the majority class using the **`step_downsample()`** function from the **`themis`** package. Then, we used 5-fold cross-validation with stratification by the **`is_antonoff`** variable to estimate the performance of the model. The model was fit using `tidymodels` [@citetidymodels].

## Random Forest

Random forests (RF) build an ensemble of decision trees during the training phase, and then use majority voting for classification tasks. For each input of data, each decision tree in the forest makes a prediction, and the final classification is determined by the majority vote of these predictions. For example, if more trees in the ensemble predict a certain class for the input data than any other class, the RF model assigns that class to the input. Each tree in the ensemble is trained on a random sample of data taken with replacement, and only a random subset of features are considered when splitting at each node. This method ensures that each tree in the ensemble is unique, avoiding overfitting and capturing non-linear relationships and more complex patterns in the data.

The training process for the RF model can be summarized as follows (@citehastie) :

>>>>

---
title: "Detecting Stylistic Variation in Pop Production"
author: 
  - name: Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/spotify_pop. We thank Rohan Alexander, Carl Wilson and Inessa De Angelis, Lindsay Katz, Michael Chong, and Ciara Zogheib for valuable suggestions. For any comments or suggestions, please contact michaela.drouillard@mail.utoronto.ca"
date: "today"
date-format: "long"
abstract: "We investigate the feasibility of predicting whether a song was produced by Jack Antonoff using Spotify's audio feature API. We develop two models: a logistic regression model and a random forest model. The logistic regression model, focusing on Antonoff's collaborators' discographies, demonstrates an accuracy of 76%, precision of 39%, and a recall of 76% when predicting whether a track has an Antonoff credit. The random forest model, which includes discographies of other notable pop producers, shows an accuracy of 61% when predicting tracks' producers, comparing favorably to a No Information Rate of 43%. A significant finding is that 'danceability', a metric developed from human-tagged data, emerges as a crucial predictor in both models. This research contributes to the broader understanding of computational analysis in cultural contexts, illustrating how a blend of human judgment and algorithmic processing can approximate musical styles."
format: pdf
number-sections: true
bibliography: references.bib
csl: modern-language-association.csl
number-depth: 2
---

@@

library(tidyverse)
library(lubridate)
library(tidyr)
library(dplyr)
library(tidymodels)
library(randomForest)
library(ranger)
library(caret)
library(themis)
library(yardstick)
library(kableExtra)
library(dplyr)
```

@@

It is widely acknowledged that style recognition in music is difficult. Style, which is hard enough to define without building computational representations of it, is often constructed using lower-level perceptual features such as a pitch and tempo. We are interested in understanding how to create an approximation for production style based on a dataset built using the Spotify audio features API. This includes both crowdsourced features and features derived from machine listening techniques. To do this, we explore different audio feature metrics and identify which are the most import predictors in identifying pop producers' style. We focus on Jack Antonoff, as he is a prolific and widely used producer, who also has his own solo acts.

@@

We obtained tracks from six artists with extensive discographies who have collaborated with Antonoff, both before and during their collaboration. We also obtained tracks from similarly prolific pop producers: Joel Little, Ariel Rechstaid, Max Martin, Greg Kurstin, Paul Epworth and Rick Nowels. Our dataset contains 3,738 tracks in total, which were split into a training and test sets.

We built two models to explore the data. The first model, a logistic regression, is employed to observe any patterns in audio feature scores across Antonoff's collaborations with artists. The visualizations and analysis offer a closer look at Antonoff's oeuvre, along with a portrait of the trends in female singer-songwriter contemporary pop (according to Spotify's API). It draws on a dataset of artists who have collaborated with Antonoff (1,954 tracks total), with a binary variable to tag whether the song was an Antonoff production. The training set included 1,465 tracks, and the testing set included 489 tracks.

The second model, a random forest, trains on a dataset that includes the discographies of other prominent pop producers to classify producers. This expands on and complements the results of the logistic regression -- where a logistic regression predicting an Antonoff-produced track in an artist's discography might be picking up on an artist releasing pop music, training a random forest on a bigger dataset with more producers captures both non-linear relationships in the data and accounts for intra-genre differences. In obtaining the strongest predictor variables in the random forest, we can better understand how it distinguishes between different pop producers' styles.

@@

In an interview with Spotify principal engineer Glen McDonald, we learned that the danceability and valence features were built using crowdsourced methods. With the understanding that what we understand as "data-driven" in recommendation systems is often intertwined with human subjectivity -- engineers' understandings of taste and taste-making also shape the recommendation systems that they build (@citeSeaver) -- our results suggest that it is possible to create representations of style by combining features derived from both automated audio analysis and more crowdsourced, subjective contributions. This research adds to a more informed understanding of how style and tastes can be algorithmically rendered using Spotify's audio features.

The remainder of this paper is structured as follows: @sec-data details the data curation and extraction processes, including documentation on features from Spotify's audio features dataset. @sec-model specifies the models used and @sec-results details their performance and most important predictors. @sec-discussion situates the study in the broader context of computational stylistics and recommendation systems, detailing how we can conceptualize "style" given the results and limitations of our study. We provide background on the Spotify features' provenance based on an interview we conducted with Glen McDonald, a principal engineer at Spotify.

@@

We collected data from the Spotify API, which we accessed using the `spotifyr` package in R [@citeSpotify; @citeR]. We first acquired audio feature data on the complete Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Van Etten, and Mitski were included to represent other contemporary pop artists with similar fan bases who have never collaborated with Antonoff. Of the artists who did collaborate with Antonoff, we chose artists who have produced multiple albums, and at least one album with Antonoff. This dataset contains 1,954 tracks in total.

For our second dataset, we acquired the discographies of comparable contemporary pop producers by scraping their respective discography Wikipedia pages. Tracks were included if producers were either the solo producer, the primary producer, or a co-producer (a limitation of this study being that we cannot know or truly parse who takes ownership for what during creative collaborations, especially when our audio features use one number to describe an entire song). The producers include Epworth, Rechstaid, Max Martin, Nowels, Joel Little, and Greg Kustin. These producers were chosen based both on their popularity, and because some have collaborated with the same artists as Antonoff. We combined these tracks with the first datasets tracks, for a total of 3,738 tracks from 828 different artists[^1].

@@

We created a binary variable, `is_antonoff`, which is 1 if Jack Antonoff produced or co-produced the song, and 0 if it was produced by somebody else. The information underpinning this discography was drawn from the "Jack Antonoff Production Discography" Wikipedia page [@citewiki].

```{r}
#| label: tbl-producercounts
#| tbl-cap: Number of tracks per producer in the original dataset
#| echo: false
#| warning: false

@@

tbl |>
  kable(
    col.names = c("Producer", "Number of Tracks"),
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```

@@

Spotify provides variables for each song, some of which they developed, which enables comparisons across artists. The final audio features in our dataset are: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `is_antonoff`. Spotify documents the `get_artist_audio_features` variables as follows (@citeSpotifyDocumentation):

> **acousticness**: A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.
>
> **danceability**: Describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.
>
> **energy**: A measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.
>
> **instrumentalness**: Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0.
>
> **speechiness:** Detects the presence of spoken words in a track. The more exclusively speech-like the recording (e.g. talk show, audio book, poetry), the closer to 1.0 the attribute value. Values above 0.66 describe tracks that are probably made entirely of spoken words. Values between 0.33 and 0.66 describe tracks that may contain both music and speech, either in sections or layered, including such cases as rap music. Values below 0.33 most likely represent music and other non-speech-like tracks.
>
> **key**: The key the track is in. Integers map to pitches using standard [Pitch Class notation](https://en.wikipedia.org/wiki/Pitch_class). E.g. 0 = C, 1 = C♯/D♭, 2 = D, and so on. If no key was detected, the value is -1.
>
> **liveness**: Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.
>
> **loudness:** The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.
>
> **mode:** Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0.
>
> **tempo:** The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.
>
> **valence:** A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).

By analyzing the cumulative distribution functions of various audio features, particularly when categorizing songs based on whether they were produced by Antonoff (@fig-cdf), distinct patterns emerge. Notably, while attributes such as liveness, instrumentalness, and speechiness exhibit relative homogeneity, there are slight disparities in features like danceability, energy, loudness, acousticness, and valence. (@fig-means in @sec-datarange depicts the range of scores for the tracks across all variables except for tempo, mode, and loudness, which are measured using different metrics.)

```{r, fig.width= 8, fig.height=6}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of each variable from the Spotify Audio Features API for our full dataset
#| echo: false
#| warning: false

av <- read_csv(here::here("inputs/data/antonoff_viz.csv"))

av$is_antonoff <- ifelse(av$producer == "antonoff", 1, 0)
av$is_antonoff <- as.factor(av$is_antonoff)
levels(av$is_antonoff) <- c("No", "Yes")

av |>
  pivot_longer(
    cols = c(
      danceability,
      energy,
      loudness,
      mode,
      speechiness,
      acousticness,
      instrumentalness,
      liveness,
      valence,
      tempo
    ),
    names_to = "name",
    values_to = "value"
  ) |>
  ggplot(aes(x = value, color = is_antonoff)) +
  stat_ecdf() +
  facet_wrap(vars(name),
             nrow = 3,
             ncol = 4,
             scales = "free") +
  theme_minimal() +
  labs(x = "Value",
       y = "Proportion",
       color = "Antonoff")
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-energy
#| fig-cap: Energy scores for contemporary pop artists over time
#| echo: false
#| warning: false

av <- av |>
  mutate(is_antonoff = factor(is_antonoff))

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-danceability
#| fig-cap: Danceability scores for contemporary pop artists over time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = danceability, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Danceability") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-valence
#| fig-cap: Valence scores for contemporary pop artists over time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = valence, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Valence") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-acousticness
#| fig-cap: Acousticness scores for contemporary pop artists over time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = acousticness, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Acousticness") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-instrumentalness
#| fig-cap: Instrumentalness scores for contemporary pop artists over time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = instrumentalness, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Instrumentalness") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

```{r, fig.width= 8, fig.height=6}
#| label: fig-tempo
#| fig-cap: Tempo scores for contemporary pop artists over time
#| echo: false
#| warning: false

av |>
  filter(album_release_date_precision == "day") |>
  mutate(album_release_date = ymd(album_release_date)) |>
  ggplot(aes(y = tempo, x = album_release_date, color = is_antonoff)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Antonoff", x = "Album Release Date", y = "Tempo") +
  facet_wrap(~ artist_name, ncol = 3)
```

@@

To be clear, we do not suggest that Antonoff's involvement is the cause for the change in artist's sound. Arguments of "Antonoffication" risk removing agency and artistic credit from (female) artists, and attributing both their critical success and failures to Antonoff (@citeWilson). In an industry primarily led by male producers, and using features so limited in their ability to describe anything about the process of creation, it would not be appropriate to use these trends to suggest that Antonoff's involvement is the cause behind changes in sound. The purpose of these visualizations is to observe the distribution of variables across artists, over time, with a focus on Antonoff.

This exploratory data analysis sets the stage for:

a) employing logistic regression to quantify the likelihood of Antonoff's involvement in a song based on these audio features, 
b) classifying tracks produced by a wider range of producers using a random forest, 
c) identifying the most influential variables in this classification to better understand how style can be inferred from the provided dataset features.

@@

The aim of the logistic regression is to determine whether we can predict Antonoff's influence on artists' sound. We select a total of 1,465 tracks from the dataset from 12 artists: Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Van Etten, Mitski, and Bleachers. Marina and the Diamonds, Maggie Rogers, Sharon Van Etten and Mitski have not collaborated with Antonoff, but their tracks were added to contribute more data from the same taste pool to provide more context for the model.

@@

$$
\begin{aligned}
\mbox{Pr}(y_i = 1) = & \text{logit}^{-1}(β_0 + β_1 \times \text{danceability} + β_2 \times \text{energy} \\
& + β_3 \times \text{loudness} + β_4 \times \text{mode} + β_5 \times \text{speechiness} \\
& + β_6 \times \text{acousticness} + β_7 \times \text{instrumentalness} \\
& + β_8 \times \text{liveness} + β_9 \times \text{valence} + β_{10} \times \text{tempo})
\end{aligned}
$$

where $\text{Pr}(y_i=1)$ is the probability of a song being produced by Antonoff, $\beta_0$ is the intercept, and $\beta_1$, $\beta_2$, $\beta_3$, $\beta_4$, $\beta_5$, $\beta_6$, $\beta_7$, $\beta_8$, $\beta_9$, and $\beta_{10}$ are the coefficients corresponding to the 10 predictor variables.

After pre-processing using the `tidyverse` package [@citetidyverse] in R [@citeR], we followed Julia Silge's downsampling methodology to build a model that accounts for imbalanced classes (@citeSilge). We downsampled the majority class using the **`step_downsample()`** function from the **`themis`** package. Then, we used 5-fold cross-validation with stratification by the **`is_antonoff`** variable to estimate the performance of the model. The model was fit using `tidymodels` [@citetidymodels].

## Random forest

Random forests build an ensemble of decision trees during the training phase, and then use majority voting for classification tasks. For each input of data, each decision tree in the forest makes a prediction, and the final classification is determined by the majority vote of these predictions. For example, if more trees in the ensemble predict a certain class for the input data than any other class, the random forest model assigns that class to the input. Each tree in the ensemble is trained on a random sample of data taken with replacement, and only a random subset of features are considered when splitting at each node. This method ensures that each tree in the ensemble is unique, avoiding overfitting and capturing non-linear relationships and more complex patterns in the data.

The training process for the random forest model can be summarized as follows (@citehastie):
